---
title: "Corina's Summer Project - Methane Data"
date: "2023-07-20"
output: 
    html_document:
      theme: cerulean 
      toc: true
      toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Abstract draft
Abstract: 

Anthropogenic climate change is negatively affecting our planet and one of the main misconceptions surrounding climate change is that carbon dioxide (CO2) is the sole contributor to global warming. However, methane (CH4) has approximately 25 times the global warming potential compared to CO2 and the methane concentration in the atmosphere is a key part of this issue. Even though methane emissions have increased by 151% since the industrial revolution, methane's role in global warming is relatively understudied. Researching methane in relation to agriculture is crucial as agricultural soil is both a methane sink and source. Fertilization’s effect on methane fluxes in soil aren’t well understood despite the importance of agriculture worldwide and the need to understand its role in climate change. I conducted a meta-analysis to identify the relationship between fertilization and CH4 flux levels in soil. The analysis found that__________. This result can be used in global methane models and to inform policies that help companies and farmers conduct best practices.    


## Initial data
```{r methane data}
#Read in and clean-up data
library(dplyr)
library(tidyr)
library(stringr)
library(readr)
library(udunits2)
current_data <- read_csv("CH4_7.28.csv",
                         trim_ws = TRUE,
                         #0       10        20        30
                         #123456789012345678901234567890123456                     
                         col_types = "ccdddcdccccdddcccddddddccddddddccccc")
current_data %>%
  filter(is.na(`Discard?`)) %>%
  select(-`Discard?`) -> good_data

good_data %>%
  group_by(CH4_flux_unit_V2) %>%
  mutate(count = length(Study_number)) %>%
  select(CH4_flux_unit_V2, count) %>%
  unique(.) %>%
  arrange(desc(count)) -> unit_counts

#still some conversions to do

good_data %>%
  group_by(Manipulation_V2) %>%
  mutate(count = length(Study_number)) %>%
  select(Manipulation_V2, count) %>%
  unique(.) %>%
  arrange(desc(count)) -> manipulation_counts

#"Fertilization + Precipitation as well as Precipitation + Fertilization

#filter out units that we aren't going to use at this time

#select just columns for unit transformation
good_data %>%
  select(Study_number, RN, CH4_annual, CH4_growingseason, CH4_monthly,
         `CH4-C_converted`, Season_converted, N, CH4_flux_unit,
         CH4_flux_unit_V2, SD_CH4_annual, SD_CH4_growingseason,
         SD_CH4_monthly) -> for_transformation

#current csv has zeros for Study # 23035 where NA's should be in Season_converted
for_transformation[for_transformation$Study_number == "23035",]$Season_converted <- NA

for_transformation %>%
  filter(! is.na(CH4_annual)) %>%
  mutate(period = "annual",
         flux = CH4_annual,
         flux_sd = SD_CH4_annual) %>%
  select(- c(CH4_annual,
             SD_CH4_annual,
             CH4_growingseason,
             SD_CH4_growingseason,
             CH4_monthly,
             SD_CH4_monthly)) -> annual

for_transformation %>%
  filter(! is.na(CH4_monthly)) %>%
  mutate(period = "monthly",
         flux = CH4_monthly,
         flux_sd = SD_CH4_monthly) %>%
  select(- c(CH4_annual,
             SD_CH4_annual,
             CH4_growingseason,
             SD_CH4_growingseason,
             CH4_monthly,
             SD_CH4_monthly))-> monthly

for_transformation %>%
  filter(! is.na(CH4_growingseason)) %>%
  mutate(period = "growing season",
         flux = CH4_growingseason,
         flux_sd = SD_CH4_growingseason) %>%
  select(- c(CH4_annual,
             SD_CH4_annual,
             CH4_growingseason,
             SD_CH4_growingseason,
             CH4_monthly,
             SD_CH4_monthly))-> growing_season

annual %>%
  bind_rows(monthly) %>%
  bind_rows(growing_season) %>%
  arrange(Study_number) %>%
  mutate(converted = ifelse(is.na(Season_converted),
                            `CH4-C_converted`,
                            Season_converted),
         converted = coalesce(converted, flux),
         cf = converted/flux,
         flux_sd = flux_sd*cf) %>%
  select(-c(`CH4-C_converted`, Season_converted)) -> combined

#all fluxes are in one column but not yet in the same unit
#line 100-101 will remove any studies with insufficient data for meta analysis
#only run if complete cases are needed
combined %>%
  filter(complete.cases(.)) -> combined

combined %>%
  select(CH4_flux_unit_V2) %>%
  unique(.)

# 1 mg/m2/h         
# 2 kg/ha/yr        
# 3 μg/m2/h         
# 4 mg/m2/d         
# 5 mg/m2/yr        
# 6 g/m2/yr         
# 7 g/ha/d          
# 8 CH4 μmol/mol  #this is the only unit that we can't convert  
# 9 kg/hm2/yr       
# 10 μg/m2/s 

#put all fluxes and their error in mg CH4 per meter squared per hour
#WARNING
#study number 23030 passes through this argument unconverted
combined %>%
  mutate(stnd_flux = ud.convert(converted, CH4_flux_unit_V2, "mg/m2/h"),
         stnd_flux_sd = ud.convert(flux_sd, CH4_flux_unit_V2, "mg/m2/h")) %>%
  select(-c(flux, flux_sd, CH4_flux_unit, CH4_flux_unit_V2,
            converted, cf))-> tidy_standard_unit_data

current_data %>%
  select(Study_number, RN, Study_midyear, Ecosystem_type,
         Manipulation_V2, Manipulation_level, Latitude,
         Longitude, Elevation, Soil_type, Soil_drainage,
         SM_value, SM_sd, SM_depth, SM_unit,
         `same timescale as flux? (T/F)`, `if False, timescale`) %>%
  right_join(tidy_standard_unit_data) -> metadata

metadata %>%
  group_by(Study_number) %>%
  filter(Manipulation_V2 == "None") %>%
  mutate(Control_SM = SM_value,
         Control_SM_sd = SM_sd,
         Control_SM_depth = SM_depth,
         Control_N = N,
         Control = stnd_flux,
         Control_sd = stnd_flux_sd) %>%
  select(-c(RN, Manipulation_V2, Manipulation_level, SM_value, SM_sd, SM_depth, N, stnd_flux, stnd_flux_sd,
            SM_unit, `same timescale as flux? (T/F)`,
            `if False, timescale`)) -> controls

metadata %>%
  group_by(Study_number) %>%
  filter(Manipulation_V2 != "None") %>%
  mutate(Manip_SM = SM_value,
         Manip_SM_sd = SM_sd,
         Manip_SM_depth = SM_depth,
         Manip_N = N,
         Manip = stnd_flux,
         Manip_sd = stnd_flux_sd) %>%
  select(-c(RN, SM_value, SM_sd, SM_depth, N, stnd_flux, stnd_flux_sd)) %>%
  left_join(controls, by = c("Study_number", "Study_midyear",
                             "Ecosystem_type", "Latitude",
                             "Longitude", "Elevation",
                             "Soil_type", "Soil_drainage",
                             "period")) -> tidy_metadata

tidy_metadata %>%
  filter(! is.na(Control)) -> data4meta
library("metafor")

```
## Meta-analysis code

```{r methane data meta-analysis code }

#Effect size for methane from all data
methane_data<- escalc(n1i = Control_N, n2i = Manip_N,
                      m1i = Control, m2i = Manip,
                      sd1i = Control_sd, sd2i = Manip_sd,
                      data = data4meta, measure = "SMD",
                      slab = Study_number, append=TRUE)
ma_model_1 <- rma(yi, vi, data = methane_data)
print(summary(ma_model_1))



#Ecosystem type (Ecosystem_type)
Eco_type<- escalc(n1i = Control_N, n2i = Manip_N,
                  m1i = Control, m2i = Manip,
                  sd1i = Control_sd, sd2i = Manip_sd,
                  data = data4meta, measure = "SMD",
                  slab = Ecosystem_type, append=TRUE)

Eco_model <- rma.mv(yi, vi, data = Eco_type,
                    random = ~ 1 | Study_number,
                    mods = ~ Ecosystem_type - 1,
                    method = "REML")

Coef_meta_Eco <- coef(summary(Eco_model))
print(summary(Eco_model))


#Manipulation effects (Manipulation_V2)
Manip_type<- escalc(n1i = Control_N, n2i = Manip_N,
                    m1i = Control, m2i = Manip,
                    sd1i = Control_sd, sd2i = Manip_sd,
                    data = data4meta, measure = "SMD",
                    slab = Manipulation_V2, append=TRUE)

Manip_model <- rma.mv(yi, vi, data = Manip_type,
                      random = ~ 1 | Study_number,
                      mods = ~ Manipulation_V2 - 1,
                      method = "REML")

Coef_meta_Manip <- coef(summary(Manip_model))
print(summary(Manip_model))
```

## I need help with manip effects in ag soil ma_model_ag
```{r}
#Manipulation effects in agriculture soil
ag_data<-filter(data4meta,Ecosystem_type=="Agriculture")
Ag_type<- escalc(n1i = Control_N, n2i = Manip_N,
                 m1i = Control, m2i = Manip,
                 sd1i = Control_sd, sd2i = Manip_sd,
                 data = ag_data, measure = "SMD",
                 slab = Ecosystem_type, append=TRUE)

ma_model_Ag <- rma.mv(yi, vi, data = Ag_type,
                      random = ~ 1 | Study_number,
                      #
                      #mods line prevented anything from running - is it necessary?
                      #  mods = ~ Ecosystem_type - 1,
                      #
                      method = "REML")

Coef_meta_Ag <- coef(summary(ma_model_Ag))
print(summary(ma_model_Ag))

```
## Plots

```{r methane graphs, echo=FALSE, fig.width=11, fig.height=9}

forest(ma_model_1, slab = methane_data$Study_number)
title(main="Effect size for methane from all data")
funnel(ma_model_1)


forest(Coef_meta_Eco$estimate,ci.lb = Coef_meta_Eco$ci.lb,
       ci.ub =  Coef_meta_Eco$ci.ub ,
       annotate = TRUE,slab=c("Agriculture","Forest","Pasture","Savanna"))
title(main="Ecosystem type")
funnel(Eco_model)



forest(Coef_meta_Manip$estimate,ci.lb = Coef_meta_Manip$ci.lb,
       ci.ub =  Coef_meta_Manip$ci.ub ,
       annotate = TRUE,slab=c("Burning","Fertilization","Fumigation","Tilled"))
title(main="Manipulation effects")
funnel(Manip_model)



forest(Coef_meta_Ag$estimate,ci.lb = Coef_meta_Ag$ci.lb,
       ci.ub = Coef_meta_Ag$ci.ub ,
       annotate = TRUE,slab=c("Agriculture"))
title(main="Manipulation effects in agriculture soil")
funnel(ma_model_Ag)
```
